---
title: "01 Bayesian Reasoning for Frequentist Researchers"
subtitle: "From significance decisions to posterior uncertainty"
format:
  html:
    toc: true
    number-sections: true
execute:
  warning: false
  message: false
---

## Learning Outcomes

By the end of this module, participants should be able to:

1. Explain Bayesian inference as updating from prior to posterior using data.
2. Distinguish a confidence interval from a credible interval.
3. State research questions as posterior probability questions.
4. Use a simple binomial example to see how priors and data combine.

## The Shift

Frequentist analysis usually asks:

> How unusual would these data, or more extreme data, be if a null model were true?

Bayesian analysis asks:

> Given the data and the model assumptions, what do we now believe about the unknown quantity?

For applied researchers, the second question is often closer to the real research question.

## Familiar Example

Suppose an education researcher pilots a new reading support program with 30 learners. After the program, 21 learners reach a defined proficiency benchmark.

The raw observed proportion is:

```{r}
successes <- 21
n <- 30
successes / n
```

A frequentist analysis might report a confidence interval for the proportion. A Bayesian analysis treats the true benchmark probability as uncertain and estimates a posterior distribution for it.

## Bayes' Rule

The central relationship is:

```text
posterior proportional to likelihood x prior
```

In symbols:

```text
p(theta | data) proportional to p(data | theta) x p(theta)
```

Where:

- `theta` is the unknown quantity;
- `p(theta)` is the prior;
- `p(data | theta)` is the likelihood;
- `p(theta | data)` is the posterior.

## A Binomial Updating Example

For a proportion, a convenient teaching model is:

```text
successes ~ Binomial(n, theta)
theta ~ Beta(a, b)
```

The beta prior is useful because it is defined between 0 and 1.

```{r}
library(tidyverse)

theta_grid <- seq(0, 1, length.out = 1000)

prior_grid <- tibble(
  theta = theta_grid,
  weak = dbeta(theta, 1, 1),
  skeptical = dbeta(theta, 4, 6),
  optimistic = dbeta(theta, 8, 3)
)

prior_grid |>
  pivot_longer(-theta, names_to = "prior", values_to = "density") |>
  ggplot(aes(theta, density, color = prior)) +
  geom_line(linewidth = 1) +
  labs(
    title = "Three possible priors for the benchmark probability",
    x = "Benchmark probability",
    y = "Density"
  )
```

After observing 21 successes out of 30, the posterior for a beta-binomial model is:

```text
theta | data ~ Beta(a + successes, b + failures)
```

```{r}
failures <- n - successes

posterior_grid <- tibble(
  theta = theta_grid,
  weak = dbeta(theta, 1 + successes, 1 + failures),
  skeptical = dbeta(theta, 4 + successes, 6 + failures),
  optimistic = dbeta(theta, 8 + successes, 3 + failures)
)

posterior_grid |>
  pivot_longer(-theta, names_to = "prior", values_to = "density") |>
  ggplot(aes(theta, density, color = prior)) +
  geom_line(linewidth = 1) +
  labs(
    title = "Posterior distributions after observing the data",
    x = "Benchmark probability",
    y = "Density"
  )
```

## Posterior Summaries

```{r}
summarise_beta <- function(a, b) {
  tibble(
    mean = a / (a + b),
    lower_95 = qbeta(0.025, a, b),
    upper_95 = qbeta(0.975, a, b),
    prob_above_60 = 1 - pbeta(0.60, a, b),
    prob_above_70 = 1 - pbeta(0.70, a, b)
  )
}

bind_rows(
  weak = summarise_beta(1 + successes, 1 + failures),
  skeptical = summarise_beta(4 + successes, 6 + failures),
  optimistic = summarise_beta(8 + successes, 3 + failures),
  .id = "prior"
)
```

## Interpretation

A Bayesian statement can say:

> Under the weak prior, the posterior mean benchmark probability is about X, with a 95% credible interval from Y to Z. The posterior probability that the benchmark probability exceeds 0.60 is P.

This is different from saying:

> If we repeated the sampling procedure many times, 95% of such intervals would contain the true parameter.

The Bayesian credible interval is a probability statement about the parameter under the model.

## Exercise 1

Change the number of successes from 21 to 18 and then to 25.

For each case:

1. Plot the posterior under the three priors.
2. Compute the posterior probability that `theta > 0.60`.
3. Write one sentence explaining how the prior matters when the sample is small.

## Exercise 2

Write the following frequentist-style question as a Bayesian-style question:

> Is the pass rate significantly above 60%?

Use this structure:

```text
What is the posterior probability that ...
```

## Takeaway

Bayesian analysis does not remove assumptions. It makes them explicit and updates them with data. The main output is a posterior distribution, which can be summarized in ways that directly answer applied research questions.

