---
title: "03 Bayesian Linear Regression"
subtitle: "Translating familiar regression into posterior inference"
format:
  html:
    toc: true
    number-sections: true
execute:
  warning: false
  message: false
---

## Learning Outcomes

By the end of this module, participants should be able to:

1. Fit a Bayesian linear regression model using `brms`.
2. Interpret posterior means, credible intervals, and posterior probabilities.
3. Compare frequentist and Bayesian regression language.
4. Report whether an effect is practically meaningful.

## Research Scenario

We estimate whether an education intervention improves endline achievement after adjusting for baseline score and socioeconomic status.

```{r}
library(tidyverse)

education <- read_csv("data/education_intervention.csv")

education |>
  select(endline_score, intervention, baseline_score, ses_index, attendance_rate) |>
  summary()
```

## Frequentist Baseline

```{r}
freq_model <- lm(
  endline_score ~ intervention + baseline_score + ses_index,
  data = education
)

summary(freq_model)
```

The frequentist output provides a point estimate, standard error, test statistic, p-value, and confidence interval. It does not directly provide the probability that the intervention effect is positive or practically meaningful.

## Bayesian Model

The Bayesian model uses the same likelihood structure:

```text
endline_score_i ~ Normal(mu_i, sigma)
mu_i = alpha + beta_1 intervention_i + beta_2 baseline_score_i + beta_3 ses_index_i
```

Priors:

```text
alpha ~ Normal(60, 15)
beta_intervention ~ Normal(0, 10)
beta_baseline ~ Normal(0.6, 0.2)
beta_ses ~ Normal(0, 4)
sigma ~ Exponential(1)
```

## Fit with brms

```{r}
#| eval: false
library(brms)

bayes_model <- brm(
  endline_score ~ intervention + baseline_score + ses_index,
  data = education,
  family = gaussian(),
  prior = c(
    prior(normal(60, 15), class = "Intercept"),
    prior(normal(0, 10), class = "b", coef = "intervention"),
    prior(normal(0.6, 0.2), class = "b", coef = "baseline_score"),
    prior(normal(0, 4), class = "b", coef = "ses_index"),
    prior(exponential(1), class = "sigma")
  ),
  backend = "cmdstanr",
  chains = 4,
  cores = 4,
  iter = 2000,
  seed = 2026
)

summary(bayes_model)
```

## Posterior Interpretation

```{r}
#| eval: false
library(tidybayes)

posterior_draws <- bayes_model |>
  spread_draws(b_intervention)

posterior_draws |>
  summarise(
    mean = mean(b_intervention),
    lower_95 = quantile(b_intervention, 0.025),
    upper_95 = quantile(b_intervention, 0.975),
    prob_positive = mean(b_intervention > 0),
    prob_practical = mean(b_intervention > 3)
  )
```

Useful claims:

- The posterior mean effect is the average effect across the posterior draws.
- The 95% credible interval contains the central 95% of posterior mass.
- `Pr(effect > 0 | data)` answers whether the intervention is likely beneficial under the model.
- `Pr(effect > 3 | data)` answers whether the effect likely exceeds a practical threshold.

## Posterior Plot

```{r}
#| eval: false
posterior_draws |>
  ggplot(aes(b_intervention)) +
  geom_density(fill = "steelblue", alpha = 0.35) +
  geom_vline(xintercept = 0, linetype = "dashed") +
  geom_vline(xintercept = 3, linetype = "dotted") +
  labs(
    title = "Posterior distribution of the intervention effect",
    x = "Effect in endline score points",
    y = "Density"
  )
```

## Posterior Predictions

```{r}
#| eval: false
new_students <- tibble(
  intervention = c(0, 1),
  baseline_score = mean(education$baseline_score),
  ses_index = 0
)

predicted <- add_epred_draws(bayes_model, newdata = new_students)

predicted |>
  group_by(intervention) |>
  median_qi(.epred)
```

This expresses the expected endline score for comparable students with and without the intervention.

## Exercise

Using the Bayesian model:

1. Estimate the posterior probability that the intervention effect is positive.
2. Estimate the posterior probability that the intervention effect is greater than 3 points.
3. Create a posterior density plot.
4. Write a results paragraph for an applied education report.

## Reporting Template

> The posterior mean intervention effect was [X] score points, with a 95% credible interval from [L] to [U]. The posterior probability that the intervention improved scores was [P1], and the probability that the effect exceeded the practical threshold of 3 points was [P2]. These results suggest [interpretation in substantive language].

## Takeaway

Bayesian regression uses familiar model structures but changes the inferential output. Instead of asking whether a coefficient is statistically significant, we summarize what the posterior distribution says about the size, direction, and practical importance of the effect.

