---
title: "04 Binary and Ordinal Models"
subtitle: "Extending Bayesian regression beyond continuous outcomes"
format:
  html:
    toc: true
    number-sections: true
execute:
  warning: false
  message: false
---

## Learning Outcomes

By the end of this module, participants should be able to:

1. Explain why binary outcomes require a different likelihood from continuous outcomes.
2. Fit a Bayesian logistic regression model using `brms`.
3. Interpret effects using predicted probabilities.
4. Understand when ordinal models are preferable to treating Likert outcomes as continuous.

## Binary Outcome Scenario

A researcher studies whether teachers adopt a new instructional practice. The outcome is binary: adopted or did not adopt.

```{r}
library(tidyverse)

adoption <- read_csv("data/teacher_adoption_survey.csv")

adoption |>
  count(adopted)
```

## Frequentist Framing

In a frequentist workflow, researchers might fit:

```{r}
freq_logit <- glm(
  adopted ~ prior_training + years_experience + school_support,
  data = adoption,
  family = binomial()
)

summary(freq_logit)
```

The coefficients are on the log-odds scale, which is difficult for many audiences.

## Bayesian Logistic Model

```text
adopted_i ~ Bernoulli(p_i)
logit(p_i) = alpha + beta_1 prior_training_i + beta_2 years_experience_i + beta_3 school_support_i
```

```{r}
#| eval: false
library(brms)

adoption_model <- brm(
  adopted ~ prior_training + years_experience + school_support,
  data = adoption,
  family = bernoulli(link = "logit"),
  prior = c(
    prior(normal(0, 1.5), class = "Intercept"),
    prior(normal(0, 1), class = "b")
  ),
  backend = "cmdstanr",
  chains = 4,
  cores = 4,
  iter = 2000,
  seed = 2026
)
```

## Interpret With Predicted Probabilities

```{r}
#| eval: false
library(tidybayes)

new_teachers <- tibble(
  prior_training = c(0, 1),
  years_experience = mean(adoption$years_experience),
  school_support = 3
)

predicted_prob <- add_epred_draws(adoption_model, newdata = new_teachers)

predicted_prob |>
  group_by(prior_training) |>
  median_qi(.epred)
```

For applied audiences, predicted probabilities are usually clearer than log-odds.

## Exercise: Binary Outcome

1. Fit the Bayesian logistic model.
2. Estimate adoption probability for teachers with and without prior training.
3. Estimate the posterior distribution of the probability difference.
4. Write a two-sentence result in plain language.

## Ordinal Outcome Scenario

Many social science outcomes are ordinal:

- agreement scales;
- engagement ratings;
- satisfaction ratings;
- perceived usefulness scales.

The dataset `student_engagement_ordinal.csv` contains engagement measured from 1 to 5.

```{r}
engagement <- read_csv("data/student_engagement_ordinal.csv")

engagement |>
  count(engagement)
```

## Cumulative Ordinal Model

For an ordinal response, a cumulative model estimates shifts across ordered categories.

```{r}
#| eval: false
ordinal_model <- brm(
  ordered(engagement) ~ intervention + baseline_motivation + teacher_feedback,
  data = engagement,
  family = cumulative(link = "logit"),
  prior = c(
    prior(normal(0, 1), class = "b")
  ),
  backend = "cmdstanr",
  chains = 4,
  cores = 4,
  iter = 2000,
  seed = 2026
)
```

## Communicating Ordinal Results

For many audiences, report predicted category probabilities.

```{r}
#| eval: false
new_students <- tibble(
  intervention = c(0, 1),
  baseline_motivation = 3,
  teacher_feedback = 3
)

predicted_categories <- posterior_epred(ordinal_model, newdata = new_students)
```

Interpretation should avoid saying that the intervention "increased engagement by 0.4 Likert points" unless a continuous approximation is deliberately chosen and justified.

## Takeaway

Bayesian models are not limited to linear regression. The same workflow applies across likelihoods: choose a model appropriate for the outcome, specify priors, fit, check, and communicate results on a scale that answers the applied question.

