---
title: "05 Bayesian Hierarchical Models"
subtitle: "Partial pooling for nested social science and education data"
format:
  html:
    toc: true
    number-sections: true
execute:
  warning: false
  message: false
---

## Learning Outcomes

By the end of this module, participants should be able to:

1. Explain complete pooling, no pooling, and partial pooling.
2. Fit a varying-intercept Bayesian hierarchical model.
3. Interpret school-level variation and shrinkage.
4. Explain why hierarchical models are useful with sparse group data.

## Research Scenario

Students are nested within schools. Scores may vary by school because of leadership, resources, context, peer composition, and other unmeasured factors.

```{r}
library(tidyverse)

education <- read_csv("data/education_intervention.csv")

education |>
  count(school_id)
```

## Three Ways to Handle Schools

### Complete Pooling

Ignore school differences:

```text
endline_score ~ intervention + baseline_score + ses_index
```

### No Pooling

Estimate each school separately:

```text
endline_score ~ school_id
```

This can be unstable when some schools have few students.

### Partial Pooling

Estimate school-level variation while sharing information across schools:

```text
endline_score ~ intervention + baseline_score + ses_index + (1 | school_id)
```

Partial pooling shrinks noisy school estimates toward the overall mean, especially for schools with less data.

## Fit the Hierarchical Model

```{r}
#| eval: false
library(brms)

hier_model <- brm(
  endline_score ~ intervention + baseline_score + ses_index + (1 | school_id),
  data = education,
  family = gaussian(),
  prior = c(
    prior(normal(60, 15), class = "Intercept"),
    prior(normal(0, 10), class = "b", coef = "intervention"),
    prior(normal(0.6, 0.2), class = "b", coef = "baseline_score"),
    prior(normal(0, 4), class = "b", coef = "ses_index"),
    prior(exponential(1), class = "sd"),
    prior(exponential(1), class = "sigma")
  ),
  backend = "cmdstanr",
  chains = 4,
  cores = 4,
  iter = 2000,
  seed = 2026
)
```

## School-Level Estimates

```{r}
#| eval: false
library(tidybayes)

school_effects <- hier_model |>
  spread_draws(r_school_id[school_id, term])

school_effects |>
  filter(term == "Intercept") |>
  group_by(school_id) |>
  median_qi(r_school_id) |>
  ggplot(aes(y = reorder(school_id, r_school_id), x = r_school_id)) +
  geom_pointinterval() +
  geom_vline(xintercept = 0, linetype = "dashed") +
  labs(
    title = "School-level deviations from the overall intercept",
    x = "School effect",
    y = "School"
  )
```

## Visualizing Partial Pooling Conceptually

```{r}
school_means <- education |>
  group_by(school_id) |>
  summarise(
    n = n(),
    raw_mean = mean(endline_score),
    .groups = "drop"
  )

overall_mean <- mean(education$endline_score)

school_means |>
  ggplot(aes(x = raw_mean, y = reorder(school_id, raw_mean), size = n)) +
  geom_point(color = "firebrick") +
  geom_vline(xintercept = overall_mean, linetype = "dashed") +
  labs(
    title = "Raw school means before partial pooling",
    x = "Raw school mean endline score",
    y = "School",
    size = "n"
  )
```

Raw school means are not the same as hierarchical posterior estimates, but this plot helps show why school sample size matters.

## Exercise

1. Fit the hierarchical model.
2. Compare the intervention effect from the non-hierarchical and hierarchical models.
3. Plot school-level effects.
4. Identify one school with high uncertainty and explain why the uncertainty is high.

## Reporting Template

> We fitted a Bayesian varying-intercept model with students nested within schools. The model estimated the intervention effect while allowing baseline achievement to vary across schools through school-level intercepts. The posterior distribution of the school-level standard deviation indicated [small/moderate/large] between-school variation after accounting for student-level predictors.

## Takeaway

Hierarchical models are one of the strongest reasons applied researchers should learn Bayesian analysis. They allow researchers to estimate group-level variation while avoiding the instability of separate estimates for every group.

