---
title: "06 Model Checking and Model Comparison"
subtitle: "How to decide whether a Bayesian model is credible"
format:
  html:
    toc: true
    number-sections: true
execute:
  warning: false
  message: false
---

## Learning Outcomes

By the end of this module, participants should be able to:

1. Check basic MCMC diagnostics.
2. Use posterior predictive checks.
3. Compare models using LOO.
4. Separate computational problems from substantive model problems.

## Two Kinds of Checking

Bayesian model checking asks two different questions:

1. Did the computation work?
2. Does the model describe the data well enough for the research purpose?

Both matter.

## Computational Diagnostics

Important diagnostics:

- R-hat close to 1.00;
- adequate effective sample size;
- no divergent transitions;
- trace plots without obvious pathologies;
- no warnings that indicate serious sampler problems.

```{r}
#| eval: false
summary(bayes_model)
plot(bayes_model)
```

Interpretation:

- R-hat above 1.01 deserves attention.
- Divergent transitions should not be ignored.
- Low effective sample size means posterior summaries may be unreliable.

## Posterior Predictive Checks

A posterior predictive check asks:

> If the model were true, would simulated data look like the observed data?

```{r}
#| eval: false
library(bayesplot)

pp_check(bayes_model, ndraws = 50)
```

Useful checks:

```{r}
#| eval: false
pp_check(bayes_model, type = "dens_overlay", ndraws = 50)
pp_check(bayes_model, type = "stat", stat = "mean")
pp_check(bayes_model, type = "stat", stat = "sd")
```

## What Poor Fit Can Mean

Poor posterior predictive fit may indicate:

- wrong likelihood;
- missing predictors;
- nonlinear relationship;
- unmodeled group structure;
- outliers or heavy tails;
- inappropriate priors;
- measurement problems.

The point is not to chase a perfect model. The point is to know whether the model is adequate for the claim being made.

## Model Comparison with LOO

LOO estimates expected out-of-sample predictive performance.

```{r}
#| eval: false
library(loo)

loo_simple <- loo(bayes_model)
loo_hier <- loo(hier_model)

loo_compare(loo_simple, loo_hier)
```

LOO answers a predictive question, not a universal truth question. A model with better predictive performance is not automatically more interpretable or theoretically preferred.

## Exercise

For a fitted model:

1. Check R-hat and effective sample size.
2. Inspect trace plots.
3. Run a posterior predictive density overlay.
4. Compare two models with LOO.
5. Write a short model-checking note.

## Model-Checking Checklist

| Check | Question | Result |
|---|---|---|
| R-hat | Are values close to 1.00? | |
| ESS | Are effective sample sizes adequate? | |
| Divergences | Are there divergent transitions? | |
| Trace plots | Do chains mix well? | |
| Posterior predictive check | Do simulated data resemble observed data? | |
| Sensitivity | Do conclusions depend heavily on a questionable prior? | |
| Substantive adequacy | Is the model good enough for the claim? | |

## Takeaway

Bayesian analysis is not complete when the model finishes running. A credible workflow includes computational diagnostics, posterior predictive checks, and sensitivity checks before making substantive claims.

